Partridge: An effective system for the automatic classification of the types of academic papers

### 2013-05-24 doc/bcspaper/article.tex
tried to fix table layout
spelling and shortening of bib entries

\abstract{Partridge is a system that enables intelligent search for
  academic papers by allowing users to query terms within
  sentences designating a particular core scientific concept (e.g. Hypothesis, Result, etc). 
  The system also automatically classifies papers according to
  article types (e.g. Review, Case Study). Here we focus on the latter
  aspect of the system. For each paper, Partridge automatically
  extracts the full paper content from PDF files, converts it to XML,
  determines sentence boundaries, automatically labels the sentences
  with core scientific concepts, and then uses a random forest model to
  classify the paper type. We show that the type of a paper can be
  reliably predicted by a model which analyses the distribution of
  core scientific concepts within the sentences of the paper. We
  discuss the appropriateness of many of the existing paper types used
  by major journals, and their corresponding distributions. Partridge
  is online and available for use, includes a browser-friendly
  bookmarklet for new paper submission, and demonstrates a range of
  possibilities for more intelligent search in the scientific
  literature. The Partidge instance can be used at
  \url{http://farnsworth.papro.org.uk}, and further information about
  the project can be found at \url{http://papro.org.uk}}

@@

The value of these systems to their users is often restricted by the small
proportion of available literature that they index, forcing researchers to use
multiple, domain specific, search engines for their queries.  In contrast,
there are also a number of interdisciplinary indexing systems and online
journals such as arXiv( \url{http://arxiv.org/}),
PloSOne(\url{http://plosone.org/}), and JSTOR (\url{http://www.jstor.org/}),
that try to incorporate wide ranges of papers from as many disciplines as
possible. The traits of these systems often complement those of their
domain-specific counterparts; they provide a comprehensive collection of
literature but insufficient filtering and indexing capabilities.

However, the document title is just one of the crucial parts of a scientific
paper's structure.  Liakata \emph{et al.} (2012) describe a system for
automatically processing and classifying sentences in a research paper
according to the core scientific concept (CoreSC) that they
describe\cite{Liakata2012}.  There are 11 CoreSCs, including {\em Hypothesis},
{\em Goal}, {\it Background}, {\em Method}, {\em Result} and {\em Conclusion}.
CoreSC labels can be allocated to all sentences in a scientific paper in order
to identify which scientific concept each sentence encapsulates.  SAPIENTA
(\url{http://www.sapientaproject.com}) is a publicly available machine learning
application which can automatically annotate all sentences in a scientific
paper with their CoreSC labels. It was trained using a corpus of physical
chemistry and biochemistry research papers whose sentences were manually
annotated using the CoreSC\cite{LIAKATA10.644} scheme.  An intelligent academic
information retrieval system can use this data to provide better filtering and
search capabilities for researchers.  Partridge implements such context-aware
keyword search, by allowing researchers to search for papers where a term
appears in sentences with a specific CoreSC label (e.g only in {\em Method}
sentences). This can be used to greatly improve both the precision with which
researchers are able to perform searches for scientific literature and the
accuracy of those searches in terms of relevance to the reader.

The type of a paper ({\em review}, {\em case study}, {\em research}, {\em
perspective}, etc) is another useful feature through which a user can narrow
down the results of a search.  The type of a paper can then be used to augment
queries.  For example, a user may search for a {\em review} paper containing
the keywords ``PCR microfluidics'', or a {\em research} paper with a {\em
hypothesis} containing the keywords ``cerevisiae'' and ``glucose''.   Such
paper types are not yet standardised by journals.  We expect the structure of a
paper to reflect its paper type.  For example, review papers would be expected
to contain a large amount of background material.  In this article, we describe
the application of machine learning (using random forests) to create predictive
models of a paper's type, using the distribution of CoreSC labels found in the
full text of the paper.

@@

\subsection*{Collection of scientific articles} Partridge allows users to
upload any paper which is free of copyright restrictions.  For the purpose of
this study we needed a large set of papers that we could label with CoreSC to
investigate how this information assists classification into paper type.  Open
Access (OA) journals provide free read access to their papers, but many do not
permit the user of articles for data mining purposes. The PLoS
journals\footnote{PLoS: \url{http://www.plos.org/}} contain large volumes of OA
literature under a permissive license that allows data mining.  They also use
the Pubmed Central markup schema, which is compatible with SAPIENTA, for papers
published through their journal. The PLoS journals advanced search offers
approximately 50 types of paper through which to restrict the search.  Many of
these paper type categories contain too few papers to be useful, others are too
specific (e.g.``Message from PLoS'') or appear rather adhoc. Indeed it is not
clear how these paper types have been identified. We chose to look at a range
of types, some of which we expect to overlap or have an unclear distinction.
These are namely: Essay, Correspondence, Case Study, Perspective,
Viewpoint, Opinion, Review, Research.

@@

\begin{tabular}{p{2.4cm}p{1.5cm}p{1.5cm}p{1.5cm}}
\hline\noalign{\smallskip}
Classes & Recall & Precision & F-measure  \\
\noalign{\smallskip}\svhline\noalign{\smallskip}
Case Study      &   24.3\verb|%|    &    24.5\verb|%|     &   24.4\verb|%| \\
Correspondence  &   {\bf  54.5\verb|%|}    &    {\bf 50.5\verb|%|}    &    {\bf 52.4\verb|%|} \\ 
Essay           &   45.0\verb|%|     &   46.2\verb|%|    &    45.6\verb|%| \\
Opinion         &   24.7\verb|%|    &    23.5\verb|%|  &      24.1\verb|%| \\   
Perspective     &   25.5\verb|%|  &      30.4\verb|%|  &       27.7\verb|%| \\
Research        &   {\bf 70.0\verb|%|}    &    {\bf 61.9\verb|%|}    &    {\bf 65.7\verb|%|} \\
Review          &  {\bf 63.1\verb|%|}     &   {\bf 62.5\verb|%|}     &   {\bf 62.8\verb|%|} \\
Viewpoint       &     14.9\verb|%|     &   15.7\verb|%|  &      15.3\verb|%| \\
\noalign{\smallskip}\hline\noalign{\smallskip}
\end{tabular}
\caption{Per-class recall, precision and F-measure micro-averaged over 10-fold cross validation, reporting the results on the held-out validation segment}

@@

A confusion matrix for the paper types in given in Table
\ref{tab:confusion}. Research papers are usually predicted to belong
to the Research class, and are sometimes confused with Case
Study. However, Case Study papers are often predicted to be Research
or Review. The ratio between Case study, Research and Review papers is
~1:2:3 so the outcome is not entirely surprising in terms of the data
size effect. We expected Research and Case Study to share similar
paper structures but perhaps the fact that Case Study is equally
confused with Research and Review suggests two distinct types of Case
Study.  We expected the classes Essay, Opinion, Perspective and
Viewpoint to be confused, as the four labels all indicate a paper
containing an author's personal thoughts on an issue, rather than
experiment-driven science. Opinion is confused with Perspective and
Viewpoint. Perspective is confused with almost all classes, except
Research. Viewpoint is a small class and mostly confused with Opinion
and Perspective. From this it would seem that the paper types
\{Opinion, Perspective, Viewpoint\} are very similar and should be
grouped together into one category. Indeed when we trained the model
on a super class consisting of the previous three categories,
performance was improved overall with the following f-measures:
OpinionSuper: 59, Research: 67.6, Review: 63.7, Essay: 44.9,
Correspondence: 46 abd Case Study 24.8.

@@

\begin{tabular}{lllllllll}
\hline\noalign{\smallskip}
      &  Case Study   &     Corresp.      &   Essay  & Opinion &     Perspective    &     Research    &     Review  & Viewpoint \\
\noalign{\smallskip}\svhline\noalign{\smallskip}
Case Study    &    26   &     2     &   11    &    2  &      5 & 26       & 25 &       10 \\
Correspondence  &      4  &      54   &     3    &    9     &   15 & 8   &     4 &       2 \\
Essay      &  8     &   5    &   90    &    2 &       39      &  5 & 46    &    5 \\
Opinion    &    4   &     7 &       2   &     23      &  27     &   10   &  3  &      17 \\
Perspective &       11  &      16    &    49       & 30      &  51 & 6      &  26    &   11 \\
Research      &  19    &    6 &       2 &       8    &    5      &  140  &   12    &    8\\
Review       & 27     &   8     &   35    &    5      &  13    &    21   &  197 &       6\\
Viewpoint    &    7  &      9   &     3     &   19     &   13 & 10     &   2  &      11 \\
\noalign{\smallskip}\hline\noalign{\smallskip}
\end{tabular}]
\vspace{-15pt}
\end{table}

@@

We inspected the detail of a single decision tree, grown on the entire dataset,
to give further insight into which CoreSC class decisions were responsible for
the paper type predictions. This was a very large tree, with a depth of 37
nodes in places.  The first decision was regarding the number of Background
sentences in the article. A low Background percentage, of less than 0.694
indicates a Correspondence paper. 28 out of the 31 papers that were classified
by this decision as Correpondence were correctly labelled.

@@

The largest node classifiying Essay did so via a route after the low
Experiment decision that asked for a Background $>$ 48\%, but then low
values for Goal, Hypothesis, Result, Observation, Model, Conclusion
and Object. These decisions seem reasonable and agreed with our
expectations of the content of an Essay.

@@

To summarise, we have demonstrated that paper type can largely be
predicted from the distribution of the automatically obtained CoreSC
sentence labels in the full text of a paper. We have described some of
the particular CoreSC features that determine a paper type
(e.g. Corresponsence characterised by very little Background, Opinions
characterised by low percentage of Experiment) and discussed which of
the paper types are not easily separable in this way. We recommended
for example merging the Opinion, Viewpoint and Perspective articles
into a single category, which increased overall
classification performance. It is also potentially useful to distinguish between two types of Case study.
Analysis of an example tree shows the decision making
process to be complex, but agrees with our general understanding of
paper types. The Partridge online system now allows refinement of
paper search using CoreSC and paper type, both of which can be
intelligently determined using machine learning methods.

>>>>

\abstract{Partridge is a system that enables intelligent search for
  academic papers by allowing users to query terms within
  sentences designating a particular core scientific concept (e.g. Hypothesis, Result, etc). 
  The system also automatically classifies papers according to
  article types (e.g. Review, Case Study). Here we focus on the latter
  aspect of the system. For each paper, Partridge automatically
  extracts the full paper content from PDF files, converts it to XML,
  determines sentence boundaries, automatically labels the sentences
  with core scientific concepts, and then uses a random forest model to
  classify the paper type. We show that the type of a paper can be
  reliably predicted by a model which analyses the distribution of
  core scientific concepts within the sentences of the paper. We
  discuss the appropriateness of many of the existing paper types used
  by major journals, and their corresponding distributions. Partridge
  is online and available for use, includes a browser-friendly
  bookmarklet for new paper submission, and demonstrates a range of
  possibilities for more intelligent search in the scientific
  literature. The Partridge instance and further information about
  the project can be found at \url{http://papro.org.uk}}

@@

The value of these systems to their users is often restricted by the small
proportion of available literature that they index, forcing researchers to use
multiple, domain specific, search engines for their queries.  In contrast,
there are also a number of interdisciplinary indexing systems and online
journals such as arXiv and 
PloSOne
that try to incorporate wide ranges of papers from as many disciplines as
possible. The traits of these systems often complement those of their
domain-specific counterparts; they provide a comprehensive collection of
literature but insufficient filtering and indexing capabilities.

However, the document title is just one of the crucial parts of a scientific
paper's structure.  Liakata \emph{et al.} (2012) describe a system for
automatically processing and classifying sentences in a research paper
according to the core scientific concept (CoreSC) that they
describe\cite{Liakata2012}.  There are 11 CoreSCs, including {\em Hypothesis},
{\em Goal}, {\it Background}, {\em Method}, {\em Result} and {\em Conclusion}.
CoreSC labels can be allocated to all sentences in a scientific paper in order
to identify which scientific concept each sentence encapsulates.  SAPIENTA
(\url{http://www.sapientaproject.com}) is a publicly available machine learning
application which can automatically annotate all sentences in a scientific
paper with their CoreSC labels. It was trained using a corpus of physical
chemistry and biochemistry research papers whose sentences were manually
annotated using the CoreSC\cite{LIAKATA10.644} scheme.  An intelligent 
information retrieval system can use this data to provide better filtering and
search capabilities for researchers.  Partridge implements such context-aware
keyword search, by allowing researchers to search for papers where a term
appears in sentences with a specific CoreSC label (e.g only in {\em Method}
sentences). This can be used to greatly improve both the precision with which
researchers are able to perform searches for scientific literature and the
accuracy of those searches in terms of relevance to the reader.

The type of a paper ({\em Review}, {\em Case Study}, {\em Research}, {\em
Perspective}, etc) is another useful feature through which a user can narrow
down the results of a search.  The type of a paper can then be used to augment
queries.  For example, a user may search for a {\em Review} paper containing
the keywords ``PCR microfluidics'', or a {\em Research} paper with a {\em
Hypothesis} containing the keywords ``cerevisiae'' and ``glucose''.   Such
paper types are not yet standardised by journals.  We expect the structure of a
paper to reflect its paper type.  For example, review papers would be expected
to contain a large amount of background material.  In this article, we describe
the application of machine learning (using random forests) to create predictive
models of a paper's type, using the distribution of CoreSC labels found in the
full text of the paper.

@@

\subsection*{Collection of scientific articles} Partridge allows users to
upload any paper which is free of copyright restrictions.  For the purpose of
this study we needed a large set of papers that we could label with CoreSC to
investigate how this information assists classification into paper type.  Open
Access (OA) journals provide free read access to their papers, but many do not
permit the user of articles for data mining purposes. The PLoS
journals\footnote{PLoS: \url{http://www.plos.org/}} contain large volumes of OA
literature under a permissive license that allows data mining.  They also use
the PubMed Central markup schema, which is compatible with SAPIENTA, for papers
published through their journal. The PLoS journals advanced search offers
approximately 50 types of paper through which to restrict the search.  Many of
these paper type categories contain too few papers to be useful, others are too
specific (e.g. {\it Message from PLoS}) or appear rather {\it ad hoc}. Indeed it is not
clear how these paper types have been identified. We chose to look at a range
of types, some of which we expect to overlap or have an unclear distinction.
These are namely: Essay, Correspondence, Case Study, Perspective,
Viewpoint, Opinion, Review, Research.

@@

\caption{Per-class recall, precision and F-measure micro-averaged over 10-fold cross validation, reporting the results on the held-out validation segment}

\begin{tabular}{p{2.4cm}p{1.5cm}p{1.5cm}p{1.5cm}}
\hline\noalign{\smallskip}
Classes & Recall & Precision & F-measure  \\
\noalign{\smallskip}\svhline\noalign{\smallskip}
Case Study      &   24.3\verb|%|    &    24.5\verb|%|     &   24.4\verb|%| \\
Correspondence  &   {\bf  54.5\verb|%|}    &    {\bf 50.5\verb|%|}    &    {\bf 52.4\verb|%|} \\ 
Essay           &   45.0\verb|%|     &   46.2\verb|%|    &    45.6\verb|%| \\
Opinion         &   24.7\verb|%|    &    23.5\verb|%|  &      24.1\verb|%| \\   
Perspective     &   25.5\verb|%|  &      30.4\verb|%|  &       27.7\verb|%| \\
Research        &   {\bf 70.0\verb|%|}    &    {\bf 61.9\verb|%|}    &    {\bf 65.7\verb|%|} \\
Review          &  {\bf 63.1\verb|%|}     &   {\bf 62.5\verb|%|}     &   {\bf 62.8\verb|%|} \\
Viewpoint       &     14.9\verb|%|     &   15.7\verb|%|  &      15.3\verb|%| \\
\noalign{\smallskip}\hline\noalign{\smallskip}
\end{tabular}

@@

A confusion matrix for the paper types in given in Table
\ref{tab:confusion}. Research papers are usually predicted to belong
to the Research class, and are sometimes confused with Case
Study. However, Case Study papers are often predicted to be Research
or Review. The ratio between Case Study, Research and Review papers is
~1:2:3 so the outcome is not entirely surprising in terms of the data
size effect. We expected Research and Case Study to share similar
paper structures but perhaps the fact that Case Study is equally
confused with Research and Review suggests two distinct types of Case
Study.  We expected the classes Essay, Opinion, Perspective and
Viewpoint to be confused, as the four labels all indicate a paper
containing an author's personal thoughts on an issue, rather than
experiment-driven science. Opinion is confused with Perspective and
Viewpoint. Perspective is confused with almost all classes, except
Research. Viewpoint is a small class and mostly confused with Opinion
and Perspective. From this it would seem that the paper types
\{Opinion, Perspective, Viewpoint\} are very similar and should be
grouped together into one category. Indeed when we trained the model
on a super class consisting of the previous three categories,
performance was improved overall with the following F-measures:
{\bf OpinionSuper: 59, Research: 67.6, Review: 63.7, Essay: 44.9,
Correspondence: 46 and Case Study 24.8}.

@@

\begin{tabular}{lllllllll}
\hline\noalign{\smallskip}
      &  Case Study   &     Corresp.      &   Essay  & Opinion &     Perspective    &     Research    &     Review  & Viewpoint \\
\noalign{\smallskip}\svhline\noalign{\smallskip}
Case Study    &    26   &     2     &   11    &    2  &      5 & 26       & 25 &       10 \\
Correspondence  &      4  &      54   &     3    &    9     &   15 & 8   &     4 &       2 \\
Essay      &  8     &   5    &   90    &    2 &       39      &  5 & 46    &    5 \\
Opinion    &    4   &     7 &       2   &     23      &  27     &   10   &  3  &      17 \\
Perspective &       11  &      16    &    49       & 30      &  51 & 6      &  26    &   11 \\
Research      &  19    &    6 &       2 &       8    &    5      &  140  &   12    &    8\\
Review       & 27     &   8     &   35    &    5      &  13    &    21   &  197 &       6\\
Viewpoint    &    7  &      9   &     3     &   19     &   13 & 10     &   2  &      11 \\
\noalign{\smallskip}\hline\noalign{\smallskip}
\end{tabular}
\vspace{-15pt}
\end{table}

@@

We inspected the detail of a single decision tree, grown on the entire dataset,
to give further insight into which CoreSC class decisions were responsible for
the paper type predictions. This was a very large tree, with a depth of 37
nodes in places.  The first decision was regarding the number of Background
sentences in the article. A low Background percentage, of less than 0.694
indicates a Correspondence paper. 28 out of the 31 papers that were classified
by this decision as Correspondence were correctly labelled.

@@

The largest node classifying Essay did so via a route after the low
Experiment decision that asked for a Background $>$ 48\%, but then low
values for Goal, Hypothesis, Result, Observation, Model, Conclusion
and Object. These decisions seem reasonable and agreed with our
expectations of the content of an Essay.

@@

To summarise, we have demonstrated that paper type can largely be
predicted from the distribution of the automatically obtained CoreSC
sentence labels in the full text of a paper. We have described some of
the particular CoreSC features that determine a paper type
(e.g. Correspondence characterised by very little Background, Opinions
characterised by low percentage of Experiment) and discussed which of
the paper types are not easily separable in this way. We recommended
for example merging the Opinion, Viewpoint and Perspective articles
into a single category, which increased overall
classification performance. It is also potentially useful to distinguish between two types of Case study.
Analysis of an example tree shows the decision making
process to be complex, but agrees with our general understanding of
paper types. The Partridge online system now allows refinement of
paper search using CoreSC and paper type, both of which can be
intelligently determined using machine learning methods.
